---
title: "Find Bias Features - Spatial Transcriptomics Data"
package: "BiasDetect"
author:
  - name: "Christine Hou"
    affiliation: Department of Biostatistics, Johns Hopkins University
    email: chris2018hou@gmail.com
  - name: "Jacqui Thompson"
    affiliation: Department of Biostatistics, Johns Hopkins University
    email: jthom338@jh.edu
  - name: "Stephanie Hicks"
    affiliation: Department of Biostatistics, Johns Hopkins University
    email: shicks19@jhu.edu
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{01 Find Bias Features - Spatial Transcriptomics Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Introduction

`BiasDetect` package was developed on an unrelated DLPFC dataset available through the [spatialLIBD](https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-022-08601-w) package. We first chose **binomial deviance model** from [scry](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-019-1861-6) as the feature selection method that can incorporate a batch variable into the model. Next, we compared the per-gene ranks and dispersion values when the model was run with and without a batch effect. Our data-driven thresholding approach is using the cutoffs based on the number of standard deviance (nSD) of deviance and rank difference metrics. Based on self-selected nSD cutoffs, the genes identified as biased can be filtered out.

The methodology details can be found in [Find Bias Feature](https://jac-thom.github.io/findBiasedFeatures/) written by Jacqui Thompson, and the documentation related codes can be found on [GitHub](https://github.com/jac-thom/findBiasedFeatures).

## Tutorial

### Installation

`BiasDetect` is a R package. Install development version from [GitHub](https://christinehou11.github.io/BiasDetect):

```{r 'install dev', eval = FALSE}
remotes::install("christinehou11/BiasDetect")
```

### Setup

Install `BiasDetect`

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("BiasDetect")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

Load required packages

```{r 'library', message=FALSE}
library("BiasDetect")
library("SummarizedExperiment")
library("SpatialExperiment")
library("ggplot2")
library("PRECAST")
library("scater")
library("ggspavis")
```

### Data

The dataset in this tutorial section is from [spatialHPC](https://github.com/LieberInstitute/spatial_hpc).

We used [nnSVG](https://www.nature.com/articles/s41467-023-39748-z) to filter out 2082 statistically significant spatially variable genes (SVGs) and saved in a new `spe` object. The filtered new `spe` object used for the vignettes is [here](https://github.com/christinehou11/BiasDetect/blob/main/data/spe.rda).

```{r data input,message=FALSE, warning=FALSE}
data(spe)
spe_svgs <- spe
SVGs <- rowData(spe_svgs)$gene_name
```

### Bias Genes Identification

The data frame created from `featureSelect()` function includes `nSD_dev` and `nSD_rank`, and we can use `biasDetect()` function to identify the bias genes based on the self-selected `nSD` integer thresholds for deviance and rank respectively. The default threshold is 5 for deviance and 5 for rank.

``` {r feature select}
batch_df <- featureSelect(spe_svgs, batch_effect = "sample_id", VGs = SVGs)
dim(batch_df)
head(batch_df)
```

What's more, `biasDetect` function allows to return the biased genes either in data frame format or in plots. The default format is in data frame.

```{r bias genes identification}
bias <- biasDetect(batch_df = batch_df, nSD_dev = 10, nSD_rank = 5)
bias
```

To see the plots, add `visual = TRUE`.

```{r visual}
biasDetect(batch_df = batch_df, nSD_dev = 10, nSD_rank = 5, visual = TRUE)
```

Finally, we can obtain a new `spe` object by removing all biased genes.

```{r new spe}
svgs_filt = setdiff(rownames(spe_svgs), bias)
length(svgs_filt)
```


### Clustering

We can apply [PRECAST](https://www.nature.com/articles/s41467-023-35947-w) to identify clusters using K = 7. As we can see, the cluster are successfully distinguished after we remove the bias genes, which proves the practicability of our method.

```{r clustering, message=FALSE, warning=FALSE}
path1 <- system.file("extdata","seuInt-hpc_k-7_svgs_metadata.csv", package="BiasDetect")
clusters1 <- read.csv(path1, row.names = 1)
dim(clusters1)
identical(rownames(clusters1), rownames(colData(spe_svgs)))
spe_svgs$precast_k7 = clusters1$cluster
spe_svgs$precast_k7_ordered = factor(
    spe_svgs$precast_k7, levels=c(7,2,3,1,6,5,4), 
    labels=c("WM","WM (2)","SR/SL","CA1","CA3","DG GCL","DG ML"))

path2 <- system.file("extdata","seuInt-hpc_k-7_svgs-no-bias_metadata.csv", package="BiasDetect")
clusters2 <- read.csv(path2, row.names = 1)
identical(rownames(clusters2), rownames(colData(spe_svgs)))
spe_svgs$precast_k7_nobias = clusters2$cluster
spe_svgs$precast_k7_nobias_ordered= factor(
    spe_svgs$precast_k7_nobias, levels=c(1,2,7,5,6,4,3),
    labels=c("WM","SR/SL","CA1","CA1 (2)","CA3","DG GCL","DG ML"))

spe_svgs <- logNormCounts(spe_svgs)
markers = c("MBP","GFAP","SPARCL1","FIBCD1","COL5A2","KCNQ5","CARTPT","PCDH8","CALB1")
plotGroupedHeatmap(spe_svgs, features = markers, swap_rownames="gene_name", 
                   group="precast_k7_ordered",
                   scale=TRUE, center=TRUE, 
                   cluster_rows=FALSE, cluster_cols=FALSE)
plotGroupedHeatmap(spe_svgs, features = markers, swap_rownames="gene_name",
                   group="precast_k7_nobias_ordered",
                   scale=TRUE, center=TRUE, 
                   cluster_rows=FALSE, cluster_cols=FALSE)

l2 = unique(spe_svgs$sample_id)
names(l2) = l2
l2 = lapply(l2, function(x) spe_svgs[,colData(spe_svgs)$sample_id==x])

col.pal1 = c("#1f77b4FF","#aec7e8FF","#ffbb78FF","#2ca02cFF","#ff7f0eFF","#d62728FF","#ff9896FF")
col.pal2 = c("#1f77b4FF","#ffbb78FF","#2ca02cFF","#98df8aFF","#ff7f0eFF","#d62728FF","#ff9896FF")

c1 <- lapply(seq_along(l2), function(x) {
  plotSpots(l2[[x]], annotate="precast_k7_ordered")+
    labs(color="clus")+
    scale_color_manual(values=col.pal1)+
    theme(plot.title=element_text(size=8))
})
drawFigs(c1, layout.dim = c(1, 4), common.legend = TRUE, legend.position = "right", align = "h")

c2 <- lapply(seq_along(l2), function(x) {
  plotSpots(l2[[x]], annotate="precast_k7_nobias_ordered")+
    labs(color="clus")+
    scale_color_manual(values=col.pal2)+
    theme(plot.title=element_text(size=8))
})
drawFigs(c2, layout.dim = c(1, 4), common.legend = TRUE, legend.position = "right", align = "h")
```

Please read [reference documentation](https://github.com/jac-thom/findBiasedFeatures) for detailed codes and scripts conducting PRECAST clustering [before](https://github.com/jac-thom/findBiasedFeatures/blob/main/code/02_spatialHPC_PRECAST-svgs.R) and [after](https://github.com/jac-thom/findBiasedFeatures/blob/main/code/05_spatialHPC_PRECAST-svgs-no-bias.R) removing the biased gene.

# `R` session information {.unnumbered}

```{r 'sessionInfo', echo=FALSE}
## Session info
options(width = 120)
sessionInfo()
```
