---
title: "Find Bias Features - Spatial Transcriptomics Data"
package: "BiasDetect"
author:
  - name: "Christine Hou"
    affiliation: Department of Biostatistics, Johns Hopkins University
    email: chris2018hou@gmail.com
  - name: "Jacqui Thompson"
    affiliation: Department of Biostatistics, Johns Hopkins University
    email: jthom338@jh.edu
  - name: "Stephanie Hicks"
    affiliation: Department of Biostatistics, Johns Hopkins University
    email: shicks19@jhu.edu
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{01 Find Bias Features - Spatial Transcriptomics Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Introduction

`BiasDetect` package was developed on an unrelated DLPFC dataset available through the [spatialLIBD](https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-022-08601-w) package. We first chose **binomial deviance model** from [scry](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-019-1861-6) as the feature selection method that can incorporate a batch variable into the model. Next, we compared the per-gene ranks and dispersion values when the model was run with and without a batch effect. Our data-driven thresholding approach is using the cutoffs based on the number of standard deviance (nSD) of deviance and rank difference metrics. Based on self-selected nSD cutoffs, the genes identified as biased can be filtered out.

The methodology details can be found in [Find Bias Feature](https://jac-thom.github.io/findBiasedFeatures/) written by Jacqui Thompson, and the documentation related codes can be found on [GitHub](https://github.com/jac-thom/findBiasedFeatures).

## Tutorial

### Installation

`BiasDetect` is a R package. Install development version from [GitHub](https://christinehou11.github.io/BiasDetect):

```{r 'install dev', eval = FALSE}
remotes::install("christinehou11/BiasDetect")
```

### Load Necessary Packages

Install additional required packages before running package codes.

```{r 'pkg install', eval=FALSE}
pkgs <- c("dplyr", "tidyr", "scran", "scry")
required_pkgs <- pkgs[!pkgs %in% rownames(installed.packages())]
BiocManager::install(required_pkgs)
```

Load additional packages. `dplyr` package is widely used in this vignettes to conduct data wrangling and specific information extraction.

```{r 'library', message=FALSE}
library(BiasDetect)
library(SummarizedExperiment)
library(ggplot2)
```

### Data

The dataset in this tutorial section is from [spatialHPC](https://github.com/LieberInstitute/spatial_hpc).

```{r data input}
data(spe)
spe_svgs <- spe
SVGs <- rowData(spe)$gene_name

batch_df <- featureSelect(spe_svgs, batch_effect = "sample_id", VGs = SVGs)
batch_df
```

### Bias Genes Identification

# `R` session information {.unnumbered}

```{r 'sessionInfo', echo=FALSE}
## Session info
options(width = 120)
sessionInfo()
```
